Towards Proof-Reconstruction of Problems in Classical Propositional Logic in Agda

### 2017-07-28 paper/main.tex
minor changes.

\section{Introduction}
Proof reconstruction is a hard labor, it integrates two
complex systems. On one hand, we have the automatic theorem provers (henceforth
ATPs) that depending on the specification logic and background theories exist
many alternatives. They are SAT solvers (e.g. \verb!zChaff!~\cite{Moskewicz2001} and \verb!MiniSat!~\cite{Een2004}) to prove unsatisfiability of CNF formulas,
QBF solvers (e.g. \verb!GhostQ!~\cite{Klieber2014} and~\verb!DepQBF!~\cite{Lonsing2017}) to prove satisfiability and invalidity of quantified Boolean
formulas, SMT solvers (e.g \verb!CVC4!~\cite{Barrett2011}, \verb!veriT!~\cite{bouton2009}, and \verb!Z3!~\cite{DeMoura2008}) to
prove unsatisfiability of formulas from first-order logic with theories,  and
finally, the ATPs to prove validity of formulas from first-order logic with
equality (e.g. \verb!E!~\cite{Schulz:AICOM-2002},~\verb!Gandalf!~\cite{Tammet1997},~\verb!leanCoP!~\cite{Otten2008},~\verb!Metis!~\cite{hurd2003first},~\verb!SPASS!~\cite{Weidenbach2009} and ~\verb!Vampire!~\cite{Riazanov1999}),
high-order logic (e.g. \verb!Leo-II!~\cite{Benzmuller2008} and \verb!Satallax!~\cite{Brown2012}), intuitonistic logic
(e.g. \verb!ileanCoP!~\cite{Otten2008},~\verb!JProver!~\cite{Schmitt2001}, and \verb!Gandalf!~\cite{Tammet1997}), among others.

For proof reconstruction, we need from the ATP tools proof objects in a consistent format to work with, that is, a full script describing
step-by-step the table of the derivation to get the actual proof. Unfortunately,
not all ATPs can deliver proofs or some useful output for reconstruct a proof.
For problems in classical propositional logic (henceforth CPL), from a list of
at least forty\footnote{ATPs available in the website  \texttt{SystemOnTPTP} from the TPTP World} ATPs, just a few were able to deliver proofs (e.g.
\verb!CVC4!~\cite{Barrett2011}, \verb!SPASS!, and
\verb!Waldmeister!~\cite{hillenbrand1997}) and a little bit less generate proofs in a file
format like TSTP~\cite{sutcliffe2004tstp} (e.g. \verb!E!~\cite{Schulz:AICOM-2002}, \verb!Metis!~\cite{hurd2003first}, \verb!Vampire!, and \verb!Z3!) or LFSC~\cite{Stump2008}.

On the other hand, we have the proof checkers, Interactive Theorem Provers
%(henceforth ITP)
or proof assistants (e.g. \verb!Agda!~\cite{agdateam}, \verb!Coq!~\cite{coqteam}, \verb!Isabelle!~\cite{paulson1994isabelle}, \verb!HOL4!~\cite{norrish2007hol}). These programs allow us to define the formal language for the problems like operators, logic variables, axioms, and theorems. They can assist us to check and validate the proof script delivered by the provers. Because the formalism in the source (the proof generated by the ATP) is not necessarily the same formalism in the target (proof reconstructed in the proof assistant), the reconstruction of a proof involves a deep understanding around the deduction process done by the ATP, for translating that deduction later into the proof assistant, the so called ``reverse engineering''.

@@

\verb!Metis!' proof terms encode natural deduction proofs. Its deduction system uses six simple inference rules and it proves conjectures by refutation.

@@

The \verb!strip! rule extracts a subgoal from a goal. The splitting process
takes a goal and recursively recolects all hypotesis required in order to prove the goal.
At the end, the final hypotesis represent the subgoals.

@@

\subsubsection*{Acknowledgments.}
We thank EAFIT University in Medell\'in, Colombia for funding support.

>>>>

\section{Introduction}
Proof reconstruction is a hard labor, it integrates two
complex systems. On one hand, we have the automatic theorem provers (henceforth
ATPs) that depending on the specification logic and background theories exist
many alternatives. They are SAT solvers (e.g. \verb!zChaff!~\cite{Moskewicz2001} and \verb!MiniSat!~\cite{Een2004}) to prove unsatisfiability of CNF formulas,
QBF solvers (e.g. \verb!GhostQ!~\cite{Klieber2014} and~\verb!DepQBF!~\cite{Lonsing2017}) to prove satisfiability and invalidity of quantified Boolean
formulas, SMT solvers (e.g \verb!CVC4!~\cite{Barrett2011}, \verb!veriT!~\cite{bouton2009}, and \verb!Z3!~\cite{DeMoura2008}) to
prove unsatisfiability of formulas from first-order logic with theories, and the ATPs to prove validity of formulas from first-order logic with
equality (e.g. \verb!E!~\cite{Schulz:AICOM-2002},~\verb!Gandalf!~\cite{Tammet1997},~\verb!leanCoP!~\cite{Otten2008},~\verb!Metis!~\cite{hurd2003first},~\verb!SPASS!~\cite{Weidenbach2009} and ~\verb!Vampire!~\cite{Riazanov1999}),
high-order logic (e.g. \verb!Leo-II!~\cite{Benzmuller2008} and
\verb!Satallax!~\cite{Brown2012}) or  intuitonistic logic
(e.g. \verb!ileanCoP!~\cite{Otten2008},~\verb!JProver!~\cite{Schmitt2001}, and \verb!Gandalf!~\cite{Tammet1997}), among others.

For proof reconstruction, what we need from the ATP tools is a proof object in a consistent format to work with, that is, a full script describing
step-by-step with exhaustive details and without ambiguities, the table of the
derivation to get the actual proof. For problems in classical propositional logic (henceforth CPL), from a list of
at least forty\footnote{ATPs available from the web service \texttt{SystemOnTPTP} of the TPTP World.} ATPs, just a few are able to deliver proofs (e.g.
\verb!CVC4!~\cite{Barrett2011}, \verb!SPASS!, and
\verb!Waldmeister!~\cite{hillenbrand1997}) and a little bit less reply with a proof in a file
format like TSTP~\cite{sutcliffe2004tstp} (e.g. \verb!E!, \verb!Metis!, \verb!Vampire!, and \verb!Z3!) or LFSC~\cite{Stump2008}.

On the other hand, we have the proof checkers, interactive theorem provers
(henceforth ITP) or proof assistants (e.g. \verb!Agda!~\cite{agdateam},
\verb!Coq!~\cite{coqteam}, \verb!Isabelle!~\cite{paulson1994isabelle},
\verb!HOL4!~\cite{norrish2007hol}) and their specification logic. The ITPs allow us to define the formal
language for the problems like operators, logic variables, axioms, and theorems.
With ITPs tools, we can check and validate proof scripts delivered by provers
with or without any guidance.
Because the logic of the source (the proof generated by the
ATP) is not necessarily the same logic in the target, the reconstruction turns
out in a ``reverse engineering'' task. Reconstructing a proof involves a deep
understanding around the deduction process done by the ATP and the specification
logic in the ITP.

@@

\verb!Metis!'' proof terms encode natural deduction proofs. Its deduction system uses six simple inference rules and it proves conjectures by refutation.

@@

The \verb!strip! rule extracts a subgoal from a goal. The splitting process
takes a goal and recursively recollects all hypothesis required in order to prove the goal.
At the end, the final hypothesis represent the subgoals.

@@

\subsubsection*{Acknowledgments.}
We thank EAFIT University of Medell\'in, Colombia for funding support.
This is part of Jonathan Prieto-Cubides's Master thesis in Applied
Mathematics, written under the supervision of Andr\'es Sicard-Ram\'irez
for the Logic and Computation Research Group at the EAFIT University.

We also acknowledge the work~\cite{Gomez-Londono2015} done by
Alejandro G\'omez-Londo\~no that deserves as the basis for our parsing module
for TSTP files. And last but not least, we gratefully acknowledge Andreas Abel and Chalmers University of Gothenburg, Sweden for inviting us to be part of the Agda Implementors’ Meeting XXV where we presented part of this paper.
